State-space PTA

### 2023-06-17 docs/manuscript/paper.tex
section 2 typos fix

% The MNRAS class isn't designed to include a table of contents, but for this document one is useful.
% I therefore have to do some kludging to make it work without masses of blank space.
\begingroup
\let\clearpage\relax
%\tableofcontents
\endgroup
\newpage
\section{Introduction}
The inspiral of supermassive black hole binaries \citep[SMBHBs;][]{Rajagopal1995,Wyithe2003,Sesana2013,Ravi2015MNRAS.447.2772R},  would result in the emission of nHz gravitational waves (GWs). Other potential GW sources in this low-frequency regime include cosmic strings \citep[e.g.][]{PTAstring} and cosmological phase transitions \citep[e.g.][]{PTAphase}. The detection of these long-wavelength GWs has necessitated the development of novel astrophysical methods, since it is practically impossible to engineer interferometric detectors with sufficiently long baselines. The foremost technique for the detection of GWs in this  low-frequency regime is a via timing an ensemble of pulsars; a pulsar timing array \citep[PTA;][]{2021hgwa.bookE...4V}. The presence  of a nHz GW will influence the propagation of the pulsar radio beacon, leaving a characteristic impression on the received pulsar timing signal. By measuring the modulation of the pulsar signal in this way, one can effectively construct a detector with a baseline on the scale of parsecs. Multiple PTA detectors have now been built over the last few decades, including the North American Nanohertz Observatory for Gravitational Waves \citep[NANOGrav,][]{2020ApJ...905L..34A}, the Parkes Pulsar Timing array \citep[PPTA][]{2020PASA...37...20K}, and the European Pulsar Timing Array \citep[EPTA,][]{2010CQGra..27h4014F}. These previously disparate efforts have now been joined in international collaboration, under the umbrella of the International Pulsar Timing Array \citep[IPTA,][]{2019MNRAS.490.4666P}, along with a number of newer PTAs such as the Indian Pulsar Timing Array Project \citep[InPTA,][]{ipta}, as MeerTime \citep{meertime2,Meertime} and the Chinese PTA \citep{Hobbs_2019}. \newline

@@

The intrinsic pulsar timing noise - i.e.  random, unmodelled variations in the pulse arrival time due to irregularities in the rotation of the star - has been identified as a key factor limiting the sensitivity of PTAs to GW signals \citep{Shannon2010,Lasky2015,Caballero2016}. This timing noise has multiple theorized causes including free precession \citep{free_precession_kerr,stairs_freeprecession}, microglitches \citep{Alessandro1995,Melatos2008,Espinoza2021}, asteroid encounters \citep{Shannon_2013,Brook_2014}, glitch recovery \citep{Johnston10,Hobbs2010glitch}, fluctuations in both the internal and external stochastic torques \citep{Cordes1981, 2006MNRAS.370L..76U,Antonelli2023}, variations in the coupling between the stellar crust and core \citep{Jones1990MNRAS.246..364J}, magnetospheric state switching \citep{magneto1,Lyne2010L} and superfluid turbulence \citep{Link10.111,Melatos2014}. In order to mitigate the influence of timing noise, PTAs are typically composed of millisecond pulsars (MSPs) which are known to be much more stable rotators with minimal timing noise. However, timing noise in MSPs could be a ``latent" phenomenon that will emerge with the increased detector sensitivity and longer timing spans necessary to detect GWs \citep{Shannon2010}. In addition to the temporally uncorrelated ``red" timing noise there are also secondary, white noise sources that must be considered such as phase jitter noise and radiometer noise \citep{Cordes2010,Lam2019,Parthasarathy2021}. \newline

@@

\section{State-Space Frequency Model}\label{sec:model}
We want to formulate the PTA analysis as a state-space problem with a separation between the intrinsic pulsar state, and the measurement state recorded by an observer. For this work we will take our state variable to be the intrinsic pulsar pulse frequency $f_{\rm p}(t)$, as measured in the momentarily comoving reference frame of an observer local to the pulsar. We will take our measurement variable to be the pulsar pulse frequency that is measured by an observer at Earth, $f_{\rm m}(t)$.  In Sections \ref{sec:psr_frequency}, \ref{sec:psr_measured} we will describe our model for $f_{\rm p}(t)$ and  $f_{\rm m}(t)$ respectively.

We will go on in Section \ref{sec:statespace} to bring together these two models into a cohesive, discretized state space model. 
\textcolor{red}{TK: Need some extra discussion here on why this choice of state/measurement space. Why not e.g. phase, TOA, voltage, etc. ?}

\subsection{Evolution of the pulsar frequency} \label{sec:psr_frequency}
We will take as our model of the intrinsic pulsar frequency $f_{\rm p}(t)$ a variation of the phenomenological model of \cite{Vargas}. Within this model, $f_{\rm p}(t)$ evolves according to a combination of both deterministic torques (i.e. electromagnetic spin-down) and stochastic torques (i.e. `spin wandering', achromatic variations in the pulse TOA intrinsic to the star). The deterministic torque is taken to arise from the pulsar magnetic dipole, with braking index $n=3$ whilst the stochastic torque is a simple white noise process. Specifically, the frequency evolves according to a Ornstein-Uhlenbeck process (equivalently a Langevin equation) with a time-dependent drift parameter:
\begin{equation}
	\frac{df_{\rm p}}{dt} = -\gamma	 [f_{\rm p} - f_{\rm EM} (t)] + \dot{f}_{\rm EM} + \xi(t)
	\label{eq:frequency_evolution}
\end{equation}
where $f_{\rm EM}$ is the solution of the electromagnetic spindown equation, $\dot{f}_{\rm EM}$ is the spin derivative, $\gamma$ a proportionality constant that specifies the mean-reversion timescale, and $\xi(t)$ a white noise process that satisfies,
\begin{equation}
	\langle \xi(t) \xi(t') \rangle = \sigma^2 \delta(t - t')
\end{equation}
for variance $\sigma^2$. For PTA analysis, we are concerned with timescales on the order of years. Consequently, we can express the EM spindown straightforwardly as
\begin{equation}
	f_{\rm EM} (t) = f_{\rm EM}(0) + \dot{f}_{\rm EM} (0)t
\end{equation}  
Completely, the frequency evolution is then given by the solution of the stochastic differential equation,
\begin{equation}
	\frac{df_{\rm p}}{dt} = -\gamma	 [f_{\rm p}- f_{\rm EM}(0) - \dot{f}_{\rm EM} (0)t] + \dot{f}_{\rm EM} (0) + \xi(t)
	\label{eq:frequency_evolution_expanded}
\end{equation}
As emphasised in \cite{Vargas}, this model for the frequency evolution is a phenomenological model that aims to qualitatively reproduce the typical behaviour of observed pulsars, rather than being derived from a physical model of the neutron star \citep[e.g. a model of the neutron star crust and superfluid, components][]{Meyers2021}. For our purposes of exploring the detection of GWs via a state space formulation it will prove sufficiently accurate and appropriate.

@@

\subsubsection{Plane GW perturbation}
We take a gravitational plane wave that perturbs a background Minkowski spacetime as
\begin{equation}
	g_{\mu \nu} = \eta_{\mu \nu} + H_{\mu \nu} e^{i(\Omega(\bar{n} \cdot \bar{x} - t) + \Phi_0)	}
\end{equation}
for Minkowski metric $ \eta_{\mu \nu}$, spatial coordinates $\bar{x}$, where the GW has a constant angular frequency $\Omega$, propagates in the $\bar{n}$-direction and has a phase offset of  $\Phi_0$. We emphasise that for this work $\Omega$ has no time dependence - we are concerned solely with monochromatic sources. We are free to choose our coordinate system such that $\Phi_0$ is the GW phase at $t=0$ at the Earth. The amplitude tensor $H_{\mu \nu}$ has zero temporal components ($H_{0 \mu} = H_{\mu 0} = 0$) whilst the spatial part is
\begin{align}
	H_{ij} = h_+ e_{ij}^+(\bar{n}) + h_{\times} e_{ij}^{\times}(\bar{n})
\end{align}
where  $h_{+,\times}$ are the polarisation amplitudes of the gravitational plane wave which are given by
\begin{align}
	h_+ &= h(1 + \cos^2 \iota) \\
		h_{\times} &= -2h\cos \iota
\end{align}
The polarisation tensors $e_{ij}^{+,\times}$ are uniquely defined by the principal axes of the wave:
\begin{align}
	e_{a b}^{+}(\hat{\Omega}) =\hat{m}_a \hat{m}_b-\hat{n}_a \hat{n}_b \\
		e_{a b}^{\times}(\hat{\Omega}) =\hat{m}_a \hat{n}_b+\hat{n}_a \hat{m}_b
\end{align}
which are in turn specified via the location of the GW source on the sky (via polar coordinate $\theta$ and azimuthal coordinate $\phi$) and the polarisation angle $\psi$
\begin{align}
	\vec{m} & =(\sin \phi \cos \psi-\sin \psi \cos \phi \cos \theta) \hat{x} \nonumber \\
	& -(\cos \phi \cos \psi+\sin \psi \sin \phi \cos \theta) \hat{y} \nonumber \\
	& +(\sin \psi \sin \theta) \hat{z} \\
	\vec{n} & =(-\sin \phi \sin \psi-\cos \psi \cos \phi \cos \theta) \hat{x} \nonumber \\
	& +(\cos \phi \sin \psi-\cos \psi \sin \phi \cos \theta) \hat{y}\nonumber  \\
	& +(\cos \psi \sin \theta) \hat{z}
\end{align}

\subsubsection{Pulse frequency as a photon}
\noindent We will consider the pulse frequency as a photon with covariant 4-momentum $p_{\mu}$. Generally, the frequency of a photon recorded by an observer with 4-velocity $u^{\mu}$ is 
\begin{equation}
	f = p_{\alpha} u^{\alpha}
	\label{eq:freq_temporal}
\end{equation}
\noindent We consider both our emitter and receiver to be stationary, such that  
\begin{equation}
	u^{\alpha}|_{\rm emitter} = u^{\alpha}|_{\rm receiver} = (1,0,0,0)
\end{equation}
\noindent Consequently the frequency can be directly identified with the temporal component of the covariant 4-momentum,
\begin{equation}
	f = p_t
\end{equation}
\noindent The expression for the evolution of the pulse frequency as measured by the observer on Earth is then,
\begin{equation}
	p_t(t_1)|_{\rm Earth} = p_t(t_0)|_{\rm source} + \int_{t = t_0}^{t=t_1} \dot{p}_t dt
	\label{eq:integral}
\end{equation}
\noindent where the overdot denotes a derivative w.r.t. $t$. Since the influence of the GW perturbation on $\dot{p}_t$ is small, we can relate the source emission and receiver times as $t_1 = t_0 + d$ and consider the photon trajectory to be an unperturbed path.

@@

>>>>

% The MNRAS class isn't designed to include a table of contents, but for this document one is useful.
% I therefore have to do some kludging to make it work without masses of blank space.
\begingroup
\let\clearpage\relax
%\tableofcontents
\endgroup
\newpage
\section{Introduction}
The inspiral of supermassive black hole binaries \citep[SMBHBs;][]{Rajagopal1995,Wyithe2003,Sesana2013,Ravi2015MNRAS.447.2772R},  is predicted to result in the emission of nHz gravitational waves (GWs). Other potential GW sources in this low-frequency regime include cosmic strings \citep[e.g.][]{PTAstring} and cosmological phase transitions \citep[e.g.][]{PTAphase}. The detection of these long-wavelength GWs has necessitated the development of novel astrophysical methods, since it is practically impossible to engineer interferometric detectors with sufficiently long baselines. The foremost technique for the detection of GWs in this  low-frequency regime is a via timing an ensemble of pulsars; a pulsar timing array \citep[PTA;][]{2021hgwa.bookE...4V}. The presence  of a nHz GW will influence the propagation of the pulsar radio beacon, leaving a characteristic impression on the received pulsar timing signal. By measuring the modulation of the pulsar signal in this way, one can effectively construct a detector with a baseline on the scale of parsecs. Multiple PTA detectors have now been built over the last few decades, including the North American Nanohertz Observatory for Gravitational Waves \citep[NANOGrav,][]{2020ApJ...905L..34A}, the Parkes Pulsar Timing array \citep[PPTA][]{2020PASA...37...20K}, and the European Pulsar Timing Array \citep[EPTA,][]{2010CQGra..27h4014F}. These previously disparate efforts have now been joined in international collaboration, under the umbrella of the International Pulsar Timing Array \citep[IPTA,][]{2019MNRAS.490.4666P}, along with a number of newer PTAs such as the Indian Pulsar Timing Array Project \citep[InPTA,][]{ipta}, as MeerTime \citep{meertime2,Meertime} and the Chinese PTA \citep{Hobbs_2019}. \newline

@@

The intrinsic pulsar timing noise -- i.e.  random, unmodelled variations in the pulse arrival time due to irregularities in the rotation of the star -- has been identified as a key factor limiting the sensitivity of PTAs to GW signals \citep{Shannon2010,Lasky2015,Caballero2016}. This timing noise has multiple theorized causes including free precession \citep{free_precession_kerr,stairs_freeprecession}, microglitches \citep{Alessandro1995,Melatos2008,Espinoza2021}, asteroid encounters \citep{Shannon_2013,Brook_2014}, glitch recovery \citep{Johnston10,Hobbs2010glitch}, fluctuations in both the internal and external stochastic torques \citep{Cordes1981, 2006MNRAS.370L..76U,Antonelli2023}, variations in the coupling between the stellar crust and core \citep{Jones1990MNRAS.246..364J}, magnetospheric state switching \citep{magneto1,Lyne2010L} and superfluid turbulence \citep{Link10.111,Melatos2014}. In order to mitigate the influence of timing noise, PTAs are typically composed of millisecond pulsars (MSPs) which are known to be much more stable rotators with minimal timing noise. However, timing noise in MSPs could be a ``latent" phenomenon that will emerge with the increased detector sensitivity and longer timing spans necessary to detect GWs \citep{Shannon2010}. In addition to the temporally uncorrelated ``red" timing noise there are also secondary, white noise sources that must be considered such as phase jitter noise and radiometer noise \citep{Cordes2010,Lam2019,Parthasarathy2021}. \newline

@@

\section{State-Space Frequency Model}\label{sec:model}
We want to formulate the PTA analysis as a state-space problem with a separation between the intrinsic pulsar state, and the measurement state recorded by an observer. For this work we will take our state variable to be the intrinsic pulsar pulse frequency $f_{\rm p}(t)$, as measured in the momentarily comoving reference frame of an observer local to the pulsar. We will take our measurement variable to be the pulsar pulse frequency that is measured by an observer at Earth, $f_{\rm m}(t)$.  
\subsection{Evolution of the pulsar frequency} \label{sec:psr_frequency}
We will take as our model of the intrinsic pulsar frequency $f_{\rm p}(t)$ a variation of the phenomenological model of \cite{Vargas}. Within this model, $f_{\rm p}(t)$ evolves according to a combination of both deterministic torques (i.e. electromagnetic spin-down) and stochastic torques (i.e. `spin wandering' or timing noise, achromatic variations in the pulse time of arrival intrinsic to the star). The deterministic torque is taken to arise from the pulsar magnetic dipole, with braking index $n=3$ whilst the stochastic torque is a simple white noise process. Specifically, the frequency evolves according to a Ornstein-Uhlenbeck process (equivalently a Langevin equation) with a time-dependent drift parameter:
\begin{equation}
	\frac{df_{\rm p}}{dt} = -\gamma	 [f_{\rm p} - f_{\rm EM} (t)] + \dot{f}_{\rm EM} + \xi(t) \ , 
	\label{eq:frequency_evolution}
\end{equation}
where $f_{\rm EM}$ is the solution of the electromagnetic spindown equation, $\dot{f}_{\rm EM}$ is the spin derivative, $\gamma$ a proportionality constant that specifies the mean-reversion timescale, and $\xi(t)$ a white noise process that satisfies:
\begin{equation}
	\langle \xi(t) \xi(t') \rangle = \sigma^2 \delta(t - t') \ ,
\end{equation}
for variance $\sigma^2$. For PTA analysis, we are concerned with timescales on the order of years. Consequently, we can express the EM spindown as a simple linear relation:
\begin{equation}
	f_{\rm EM} (t) = f_{\rm EM}(0) + \dot{f}_{\rm EM} (0)t \ .
\end{equation}  
Completely, the frequency evolution is then given by the solution of the stochastic differential equation,
\begin{equation}
	\frac{df_{\rm p}}{dt} = -\gamma	 [f_{\rm p}- f_{\rm EM}(0) - \dot{f}_{\rm EM} (0)t] + \dot{f}_{\rm EM} (0) + \xi(t)
	\label{eq:frequency_evolution_expanded}
\end{equation}
As emphasised in \cite{Vargas}, this model for the frequency evolution is a phenomenological model that aims to qualitatively reproduce the typical behaviour of observed pulsars, rather than being derived from a physical model of the neutron star \citep[e.g. a model of the neutron star crust and superfluid components][]{Meyers2021}. For our purposes of exploring the detection of GWs via a state space formulation it will prove sufficiently accurate and appropriate.

@@

\subsubsection{Plane GW perturbation}
We take a gravitational plane wave that perturbs a background Minkowski spacetime as
\begin{equation}
	g_{\mu \nu} = \eta_{\mu \nu} + H_{\mu \nu} e^{i(\Omega(\hat{n} \cdot \bar{x} - t) + \Phi_0)} \ ,
\end{equation}
for Minkowski metric $ \eta_{\mu \nu}$ with spatial coordinates $\bar{x}$, where the GW has a constant angular frequency $\Omega$, propagates in the $\hat{n}$-direction and has a phase offset of  $\Phi_0$. We emphasise that for this work $\Omega$ has no time dependence -- we are concerned solely with monochromatic sources. We are free to choose our coordinate system such that $\Phi_0$ is the GW phase at $t=0$ at the Earth. The amplitude tensor $H_{\mu \nu}$ has zero temporal components ($H_{0 \mu} = H_{\mu 0} = 0$) whilst the spatial part is
\begin{align}
	H_{ij} = h_+ e_{ij}^+(\hat{n}) + h_{\times} e_{ij}^{\times}(\hat{n}) \ ,
\end{align}
where $h_{+,\times}$ are the polarisation amplitudes of the gravitational plane wave. The polarisation tensors $e_{ij}^{+,\times}$ are uniquely defined by the principal axes $\hat{k}$, $\hat{l}$ of the wave:
\begin{align}
	e_{i j}^{+}(\hat{n}) =\hat{k}_i \hat{k}_j-\hat{l}_i \hat{l}_j \ , \\
		e_{i j}^{\times}(\hat{n}) =\hat{k}_i \hat{l}_j+\hat{l}_i \hat{k}_j \ .
\end{align}
The principal axes are in turn specified via the location of the GW source on the sky (via polar coordinate $\theta$ and azimuthal coordinate $\phi$) and the polarisation angle $\psi$
\begin{align}
	\hat{k} & =(\sin \phi \cos \psi-\sin \psi \cos \phi \cos \theta) \hat{x} \nonumber \\
	& -(\cos \phi \cos \psi+\sin \psi \sin \phi \cos \theta) \hat{y} \nonumber \\
	& +(\sin \psi \sin \theta) \hat{z} \\
	\hat{l} & =(-\sin \phi \sin \psi-\cos \psi \cos \phi \cos \theta) \hat{x} \nonumber \\
	& +(\cos \phi \sin \psi-\cos \psi \sin \phi \cos \theta) \hat{y}\nonumber  \\
	& +(\cos \psi \sin \theta) \hat{z}
\end{align}
where e.g. $\hat{x}$ is a unit vector in the direction of the $x$-axis of the reference frame. The direction of GW propagation is related to the principal axes as,
\begin{equation}
	\hat{n} = \hat{k} \times \hat{l} \ . 
\end{equation}

\subsubsection{Pulse frequency as a photon}
In general radio pulses from a pulsar are transmitted as an amplitude-modulated, radio-frequency carrier wave, where the carrier wave (i.e. the photon) is at typical radio frequencies of $\sim$ GHz whilst the pulse frequency $f_p(t)$ is $\sim $kHz. We are interested in measuring how the pulse frequency is perturbed by the presence of a GW. We define the geometric object
\begin{equation}
\vec{p} = \omega (1, -\bar{q})
\end{equation} 
where $\omega$ is the angular pulse frequency $(= 2 \pi f_p)$ and $\bar{q}$ is the local propagation direction of the radio photon. We can identify $\vec{p}$ as the 4-momentum of the radio photon. Generally the measured frequency of a photon recorded by an observer who is travelling with 4-velocity $u^{\mu}$ is given by the coordinate-independent expression:
\begin{equation}
	f = p_{\alpha} u^{\alpha} \ . 
	\label{eq:freq_temporal}
\end{equation}
We want to relate the pulse frequencies measured in two reference frames; one at the emitter, i.e. the pulsar, and one at the observer on Earth. For pulsar observations the kinematical Doppler shifts relative to the Solar System barycentre (SSB) are corrected in the generation of a timing solution so that both the pulsar and the Earth are at rest in the global Minkowski coordinates. After these kinematical corrections we can consider both our emitter and receiver to be stationary, such that  
\begin{equation}
	u^{\alpha}|_{\rm emitter} = u^{\alpha}|_{\rm receiver} = (1,0,0,0)
\end{equation}
\noindent Consequently the measured frequency in both frames can be directly identified with the temporal component of the covariant 4-momentum,
\begin{equation}
	f = p_t
\end{equation}
\noindent The evolution of the pulse frequency as measured by the observer on Earth is then,
\begin{equation}
	p_t(\tau)|_{\rm Earth} = p_t(t_0)|_{\rm source} + \int_{t = t_0}^{t=\tau} \dot{p}_t dt
	\label{eq:integral}
\end{equation}
\noindent where we use $\tau$ to denote the proper time recorded by the Earth observers clock, and the overdot denotes a derivative w.r.t. coordinate time along the geodesic $t$. Since the influence of the GW perturbation on $\dot{p}_t$ is small, we can relate the source emission and receiver times as $t_1 = t_0 + d$ and consider the photon trajectory to be an unperturbed path.

@@

wave which are given by
\begin{align}
	h_+ &= h(1 + \cos^2 \iota) \\
	h_{\times} &= -2h\cos \iota
\end{align}
